---
title: "Lab 8: Categorical Indicators" 
subtitle: "SEM in R using lavaan"
author: "Kyle M. Lang"
date: "Updated: `r format(Sys.time(), format = '%Y-%m-%d')`"
params:
  answers: true
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    df_print: paged
    css: "../../resources/style.css"
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(dplyr)
library(magrittr)
library(ggplot2)

source(here::here("code", "supportFunctions.R"))

figDir <- "../figures/"

set.seed(235711)

## Define an asis engine that will evaluate inline code within an asis block:
knit_engines$set(asis = function(options) {
  if(options$echo && options$eval) knit_child(text = options$code)
}
)

opts_chunk$set(
  include = params$answers, 
  echo = params$answer, 
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  comment = NA,
  root.dir = here::here("./")
)
```

<!-- 
Define some hacky LaTeX commands to force nice spacing between lines
NOTE: These must be called within a math environment (e.g., $\va$)
-->
\newcommand{\va}{\\[12pt]}
\newcommand{\vb}{\\[6pt]}
\newcommand{\vc}{\\[3pt]}
\newcommand{\vx}[1]{\\[#1pt]}

---

In this lab, you'll explore methods of doing CFA/SEM with categorical data using **lavaan**.

---

# Data

---

**Observing**

- `Q1`: I notice changes in my body, such as whether my breathing slows down or speeds up. 
- `Q5`: I pay attention to whether my muscles are tense or relaxed. 
- `Q9`: When I'm walking, I deliberately notice the sensations of my body moving. 
- `Q13`: When I take a shower or bath, I stay alert to the sensations of water on my body. 
- `Q17`: I notice how foods and drinks affect my thoughts, bodily sensations, and emotions. 
- `Q25`: I pay attention to sounds, such as clocks ticking, birds chirping, or cars passing. 
- `Q29`: I notice the smells and aromas of things. 
- `Q30`: I intentionally stay aware of my feelings. 
- `Q33`: I notice visual elements in art or nature, such as colors, shapes, textures, or patterns of light and shadow. 
- `Q37`: I pay attention to how my emotions affect my thoughts and behavior. 
- `Q39`: I notice when my moods begin to change.

**Describing**

- `Q2`: I'm good at finding the words to describe my feelings. 
- `Q6`: I can easily put my beliefs, opinions, and expectations into words. 
- `Q10`: I'm good at thinking of words to express my perceptions, such as how things taste, smell, or sound. 
- `Q14`: It's hard for me to find the words to describe what I'm thinking. 
- `Q18`: I have trouble thinking of the right words to express how I feel about things. 
- `Q22`: When I have a sensation in my body, it's difficult for me to describe it because I cant find the right words. 
- `Q26`: Even when I'm feeling terribly upset, I can find a way to put it into words. 
- `Q34`: My natural tendency is to put my experiences into words. 

**Acting**

- `Q3`: When I do things, my mind wanders off and I'm easily distracted. 
- `Q7`: When I'm doing something, I'm only focused on what I'm doing, nothing else. 
- `Q11`: I drive on "automatic pilot" without paying attention to what I'm doing. 
- `Q15`: When I'm reading, I focus all my attention on what I'm reading. 
- `Q19`: When I do things, I get totally wrapped up in them and don't think about anything else. 
- `Q23`: I don't pay attention to what I'm doing because I'm daydreaming, worrying, or otherwise distracted. 
- `Q27`: When I'm doing chores, such as cleaning or laundry, I tend to daydream or think of other things. 
- `Q31`: I tend to do several things at once rather than focusing on one thing at a time. 
- `Q35`: When I'm working on something, part of my mind is occupied with other topics, such as what I'll be doing later, or things I'd rather be doing. 
- `Q38`: I get completely absorbed in what I'm doing, so that all my attention is focused on it. 

**Accepting**

- `Q4`: I criticize myself for having irrational or inappropriate emotions. 
- `Q8`: I tend to evaluate whether my perceptions are right or wrong. 
- `Q12`: I tell myself that I shouldn't be feeling the way I'm feeling. 
- `Q16`: I believe some of my thoughts are abnormal or bad and I shouldn't think that way.  
- `Q20`: I make judgments about whether my thoughts are good or bad. 
- `Q24`: I tend to make judgments about how worthwhile or worthless my experiences are. 
- `Q28`: I tell myself that I shouldn't be thinking the way I'm thinking. 
- `Q32`: I think some of my emotions are bad or inappropriate and I shouldn't feel them. 
- `Q36`: I disapprove of myself when I have irrational ideas. 

---

First, we'll read in the *kims.rds* dataset, and save the resulting data frame as `kims`.

```{r}
dataDir <- "data"
kims    <- readRDS(here::here(dataDir, "kims.rds"))
```

---

# Categorical Variable CFA

---

Let's assume we want to fit the CFA model depicted in Figure \@ref(fig:cfa2). The following **lavaan** model syntax will
suffice to define our model.

```{r}
## Create separate strings for each constructs model, so I can mix and match:
obsMod <- 'obs =~ Q1 + Q5 + Q9  + Q13 + Q17 + Q25 + Q29 + Q30 + Q33 + Q37 + Q39'
desMod <- 'des =~ Q2 + Q6 + Q10 + Q14 + Q18 + Q22 + Q26 + Q34'
actMod <- 'act =~ Q3 + Q7 + Q11 + Q15 + Q19 + Q23 + Q27 + Q31 + Q35 + Q38'
accMod <- 'acc =~ Q5 + Q8 + Q12 + Q16 + Q20 + Q24 + Q28 + Q32 + Q36'

# kimsKeys <- list(
#   obs = c("Q5", "Q9", "Q13", "Q17", "Q25", "Q29", "Q30", "Q33", "Q37", "Q39"),
#   des = c("Q6", "Q10", "-Q14", "-Q18", "-Q22", "Q26", "Q34"),
#   acc = c("Q8", "Q12", "Q16", "Q20", "Q24", "Q28", "Q32", "Q36"),
#   act = c("Q7", "-Q11", "Q15", "Q19", "-Q23", "-Q27", "-Q31", "-Q35", "Q38")
# )
# 
# tmp <- scoreItems(keys = kimsKeys, items = kims)
# tmp

## Paste the construct-specific strings into a single CFA syntax string:
# cfaMod <- paste(obsMod, desMod, accMod, actMod, sep = "\n")
cfaMod <- paste(obsMod, desMod, accMod, sep = "\n")
```

---

## Robust DWLS Fit

Finally, we'll estimating the model with DWLS using robust SEs and scaled fit statistics.

```{r}
library(lavaan)

## Robust DWLS fit:
fit1.2 <- cfa(cfaMod,
              data = kims,
              std.lv = TRUE,
              ordered = TRUE,
              estimator = "WLSMV")

summary(fit1.2)
fitMeasures(fit1.2)
```

---

```{r}
## Define a few character vectors to specify which fit indices we want:
s1 <- c("chisq", "df", "pvalue")
s2 <- c("cfi", 
        "tli", 
        "rmsea", 
        "rmsea.ci.lower", 
        "rmsea.ci.upper", 
        "srmr")
s3 <- c(paste(s1, "scaled", sep = "."),
        paste(s2, "robust", sep = ".")
        )

s4 <- c(paste(c(s1, s2), "scaled", sep = "."))

## Fit measures for the robust DWLS fit:
fitMeasures(fit1.2, s4)
```

---

# SEM

---

```{r}
semMod <- paste(
  cfaMod,
  "acc ~ des + obs + age",
  "age ~~ des + obs",
  sep = "\n"
)

cats <- grep("^Q\\d", colnames(kims), value = TRUE)
cats <- setdiff(colnames(kims), "age")

semFit <- sem(semMod, data = kims, ordered = cats, estimator = "WLSMV", std.lv = TRUE)

summary(semFit)
fitMeasures(semFit)
```

---

# Measurement Invariance

---

## Configural Invariance

First, we estimate the configural invariance model by adding the grouping factor without placing any cross-group
constraints (i.e., in the same way that we did for continuous indicator models). 

```{r}
## Estimate the baseline (configurally invariant) model:
baseOut <- cfa(model            = cfaMod,
               data             = kims,
               group            = "sex",
               ordered          = TRUE,
               estimator        = "WLSMV",
               parameterization = "delta",
               std.lv           = TRUE)
```

Check the summary to ensure that everything looks sensible.

```{r}
summary(baseOut)
```

Check the model fit to test configural invariance.

```{r}
s4 <- paste(c(s1, s2), "scaled", sep = ".")

## Check the model fit:
fitMeasures(baseOut, s4)
```

The model fits well. Configural invariance holds.

---

## Threshold Invariance

Now, we need to test threshold invariance. We do so via the same procedure we would use to test invariance of factor
loadings or intercepts.

1. Estimate a restricted model wherein the thresholds are constrained to equality across groups.
1. Test the tenability of the constraint by comparing to the configurally invariant model.

We must establish threshold invariance before considering weak or strong invariance because weak and strong invariance
both concern paths linking the LRVs to the common latent factors. We must first demonstrate, therefore, that the LRVs
are defined equivalently in all groups.

---

We can make our lives much easier by using the `measEq.syntax()` function from the **semTools** package to
programmatically generate the **lavaan** model syntax for our various restricted models.

```{r}
library(semTools)

## Define the model syntax for the threshold-invariant model:
tMod <- measEq.syntax(configural.model = cfaMod,
                      data             = kims,
                      ordered          = TRUE,
                      parameterization = "delta",
                      ID.fac           = "std.lv",
                      ID.cat           = "Wu.Estabrook.2016",
                      group            = "sex",
                      group.equal      = "thresholds")
```

We can view the resulting syntax by casting the `tMod` object as a character vector and printing it with the `cat()`
function.

```{r}
tMod %>% as.character() %>% cat()
```

---

Now, we estimate the threshold invariant model. 

*Notes:*

- We do not need to specify anything for the `group.equal` argument because all invariance constraints are defined in
the model syntax we generated via `measEq.syntax()`.
- We must type cast the object returned by `measEq.syntax()` as a character vector (i.e., by wrapping it with
`as.character()`) to get the `cfa()` function to recognize this object as model syntax.

```{r}
## Estimate the threshold-invariant model:
tOut <- cfa(model     = as.character(tMod),
            data      = kims,
            group     = "sex",
            ordered   = TRUE,
            estimator = "WLSMV")

## Summarize the results:
summary(tOut)
```

As expected, the thresholds are equal across groups. Additionally, constraining the thresholds helps identify the model,
so we can freely estimate the item intercepts and the scales of the latent response variables in the second and third
groups.

```{r}
## Check the model fit:
fitMeasures(tOut, s4)
```

The model still fits very well.

---

We test threshold invariance by comparing to the configurally invariant model.

```{r}
## Test threshold invariance via model comparisons:
compareFit(baseOut, tOut) %>% summary()
```

The $\Delta \chi^2$ test is nonsignificant, so we can infer invariant thresholds.

---

## Weak Invariance

After establishing threshold invariance, the next step is testing the additional constraint of loading invariance (i.e.,
weak invariance or metric invariance).

- Again, we can use the `measEq.syntax()` function to streamline the process.

```{r}
## Define the model syntax for the threshold/loading-invariant model:
tlMod <- measEq.syntax(configural.model = cfaMod,
                       data             = kims,
                       ordered          = TRUE,
                       parameterization = "delta",
                       ID.fac           = "std.lv",
                       ID.cat           = "Wu.Estabrook.2016",
                       group            = "sex",
                       group.equal      = c("thresholds", "loadings")
                       )

## Check the resulting syntax:
tlMod %>% as.character() %>% cat()
```

---

We estimate the model in the usual way.

```{r}
## Estimate the threshold/loading-invariant model:
tlOut <- cfa(model     = as.character(tlMod),
             data      = kims,
             group     = "sex",
             ordered   = TRUE,
             estimator = "WLSMV")

## Summarize the results:
summary(tlOut)
```

Now that we've equated the factor loadings across groups, we have the additional identification necessary to freely
estimate the latent variances in the second and third groups.

```{r}
## Check the model fit:
fitMeasures(tlOut, s4)
```

Again, model fit is still excellent.

---

We evaluate weak invariance by comparing back to the threshold-invariance model.

```{r}
## Test threshold/loading invariance via model comparisons:
compareFit(tOut, tlOut) %>% summary()
```

Relative to the threshold-invariant model, we haven't lost a significant amount of fit by equating the factor loadings
across groups. Conditioning on the dubious threshold invariance, we can claim weak invariance.

---

## Strong Invariance

We conclude by equating the item intercepts across groups to test strong/metric invariance.

```{r}
## Define the model syntax for the threshold/loading/intercept-invariant model:
tliMod <-
    measEq.syntax(configural.model = cfaMod,
                  data             = kims,
                  ordered          = TRUE,
                  parameterization = "delta",
                  ID.fac           = "std.lv",
                  ID.cat           = "Wu.Estabrook.2016",
                  group            = "sex",
                  group.equal      = c("thresholds", "loadings", "intercepts")
                  )

## Check the syntax:
tliMod %>% as.character() %>% cat()

## Estimate the threshold/loading/intercept-invariant model:
tliOut <- cfa(model     = as.character(tliMod),
              data      = kims,
              group     = "sex",
              ordered   = TRUE,
              estimator = "WLSMV")

## Summarize the results:
summary(tliOut)
```

Constraining the item intercepts allows us to freely estimate the latent means in the second and third groups.

```{r}
## Check the model fit:
fitMeasures(tliOut, s4)
```

The strongly invariant model still fits very well.

```{r}
## Test threshold/loading/intercept invariance via model comparisons:
compareFit(tlOut, tliOut) %>% summary()
```

Now, we lose a significant amount of fit by constraining the item intercepts in addition to the factor loadings and
thresholds (i.e., the $\Delta \chi^2$ test is significant). So, we should reject the hypothesis of strong invariance.

---

# Moderated SEM

---

```{r}
semMod <- paste(
  cfaMod,
  'acc ~ c("b1f", "b1m") * des + c("b2f", "b2m") * obs + c("b3f", "b3m") * age',
  'age ~~ des + obs',
  sep = "\n"
)

modOut <- sem(
  model = semMod,
  data = kims,
  group = "sex",
  ordered = cats,
  estimator = "WLSMV",
  parameterization = "delta",
  std.lv = TRUE,
  group.equal = c("thresholds", "loadings")
)

summary(modOut)
```

```{r}
cons <- c('b1f == b1m', 'b2f == b2m', 'b3f == b3m')

lavTestWald(modOut, constraints = cons)
```

---

End of Lab 8

---
