---
title: "Lab 8: Categorical Indicators" 
subtitle: "SEM in R with lavaan"
author: "Kyle M. Lang"
date: "Updated: `r format(Sys.time(), format = '%Y-%m-%d')`"
params:
  answers: true
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    df_print: paged
    css: "../../resources/style.css"
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(dplyr)
library(magrittr)
library(ggplot2)

source("../../code/supportFunctions.R")

figDir <- "../figures/"

set.seed(235711)

## Define an asis engine that will evaluate inline code within an asis block:
knit_engines$set(asis = function(options) {
  if(options$echo && options$eval) knit_child(text = options$code)
}
)

opts_chunk$set(include = params$answers, 
               echo = params$answer, 
               message = FALSE,
               warning = FALSE,
               fig.align = "center",
               comment = NA)
```

<!-- 
Define some hacky LaTeX commands to force nice spacing between lines    
NOTE: These must be called within a math environment (e.g., $\va$)
-->
\newcommand{\va}{\\[12pt]}
\newcommand{\vb}{\\[6pt]}
\newcommand{\vc}{\\[3pt]}
\newcommand{\vx}[1]{\\[#1pt]}

---

In this lab, you'll explore methods of doing CFA/SEM with categorical data
using **lavaan**.

---

# Data

---

```{r, echo = FALSE}
dataDir <- "../../data/"
outlook <- readRDS(paste0(dataDir, "outlook.rds"))
```


- `Q1`: I notice changes in my body, such as whether my breathing slows down or speeds up. 
- `Q2`: I'm good at finding the words to describe my feelings. 
- `Q3`: When I do things, my mind wanders off and I'm easily distracted. 
- `Q4`: I criticize myself for having irrational or inappropriate emotions. 
- `Q5`: I pay attention to whether my muscles are tense or relaxed. 
- `Q6`: I can easily put my beliefs, opinions, and expectations into words. 
- `Q7`: When I'm doing something, I'm only focused on what I'm doing, nothing else. 
- `Q8`: I tend to evaluate whether my perceptions are right or wrong. 
- `Q9`: When I'm walking, I deliberately notice the sensations of my body moving. 
- `Q10`: I'm good at thinking of words to express my perceptions, such as how things taste, smell, or sound. 
- `Q11`: I drive on "automatic pilot" without paying attention to what I'm doing. 
- `Q12`: I tell myself that I shouldn't be feeling the way I'm feeling. 
- `Q13`: When I take a shower or bath, I stay alert to the sensations of water on my body. 
- `Q14`: It's hard for me to find the words to describe what I'm thinking. 
- `Q15`: When I'm reading, I focus all my attention on what I'm reading. 
- `Q16`: I believe some of my thoughts are abnormal or bad and I shouldn't think that way.  
- `Q17`: I notice how foods and drinks affect my thoughts, bodily sensations, and emotions. 
- `Q18`: I have trouble thinking of the right words to express how I feel about things. 
- `Q19`: When I do things, I get totally wrapped up in them and don't think about anything else. 
- `Q20`: I make judgments about whether my thoughts are good or bad. 
- `Q21`: I pay attention to sensations, such as the wind in my hair or sun on my face
- `Q22`: When I have a sensation in my body, it's difficult for me to describe it because I cant find the right words. 
- `Q23`: I don't pay attention to what I'm doing because I'm daydreaming, worrying, or otherwise distracted. 
- `Q24`: I tend to make judgments about how worthwhile or worthless my experiences are. 
- `Q25`: I pay attention to sounds, such as clocks ticking, birds chirping, or cars passing. 
- `Q26`: Even when I'm feeling terribly upset, I can find a way to put it into words. 
- `Q27`: When I'm doing chores, such as cleaning or laundry, I tend to daydream or think of other things. 
- `Q28`: I tell myself that I shouldn't be thinking the way I'm thinking. 
- `Q29`: I notice the smells and aromas of things. 
- `Q30`: I intentionally stay aware of my feelings. 
- `Q31`: I tend to do several things at once rather than focusing on one thing at a time. 
- `Q32`: I think some of my emotions are bad or inappropriate and I shouldn't feel them. 
- `Q33`: I notice visual elements in art or nature, such as colors, shapes, textures, or patterns of light and shadow. 
- `Q34`: My natural tendency is to put my experiences into words. 
- `Q35`: When I'm working on something, part of my mind is occupied with other topics, such as what I'll be doing later, or things I'd rather be doing. 
- `Q36`: I disapprove of myself when I have irrational ideas. 
- `Q37`: I pay attention to how my emotions affect my thoughts and behavior. 
- `Q38`: I get completely absorbed in what I'm doing, so that all my attention is focused on it. 
- `Q39`: I notice when my moods begin to change.


We will revisit the synthetic [*Outlook on Life Survey*][outlook0] data 
from Lab 5. Recall that the original data were collected in the United States in 
2012 to measure, among other things, attitudes about racial issues, opinions of 
the Federal government, and beliefs about the future.

We will again work with a synthesized subset of the original data. You can access
these synthetic data as [*outlook.rds*][outlook1]. This dataset comprises 
`r nrow(outlook)` observations of the following `r ncol(outlook)` variables.

- `d1:d3`: Three observed indicators of a construct measuring disillusionment 
with the US Federal government.
   - Higher scores indicate more disillusionment
   $\vb$
- `s1:s4`: Four observed indicators of a construct measuring the perceived 
achievability of material success.
   - Higher scores indicate greater perceived achievability
   $\vb$
- `progress`: A single item assessing perceived progress toward achieving the 
"American Dream"
   - Higher scores indicate greater perceived progress
   $\vb$
- `merit`: A single item assessing endorsement of the meritocratic ideal that 
hard work leads to success.
   - Higher scores indicate stronger endorsement of the meritocratic ideal
   $\vb$
- `lib2Con`: A single item assessing liberal-to-conservative orientation
   - Lower scores are more liberal, higher scores are more conservative
   $\vb$
- `party`: A four-level factor indicating self-reported political party affiliation
   $\vb$
- `disillusion`: A scale score representing disillusionment with the US Federal 
government
   - Created as the mean of `d1:d3`
   $\vb$
- `success`: A scale score representing the perceived achievability of material 
success
   - Created as the mean of `s1:s4`

---

First, we'll read in the *kims.rds* dataset, and save the resulting data frame
as `kims`.

```{r, eval = FALSE}
#dataDir <- "../data/"
dataDir <- "data/"
kims    <- readRDS(paste0(dataDir, "kims.rds"))
```

---

Now, we'll explore our data a bit and summarize the distributions of the 
*disillusionment* and *success* items.

```{r}
tmp <- outlook %>% select(d1:s4)

head(tmp)
summary(tmp)
lapply(tmp, unique)
```

Clearly, these variables are not continuous and have some sort of discrete scale. 
Although we cannot tell simply from the numeric summaries above, these items 
represent questionnaire responses on 4- and 5-point rating scales.

It would probably not be a very good idea to naively analyze these items as
continuous indicators. Let's see how we get on with a more principled treatment.

---

# Categorical Variable CFA

---

Let's assume we want to fit the CFA model depicted in Figure \@ref(fig:cfa2). 
The following **lavaan** model syntax will suffice to define our model.

```{r}
## Create separate strings for each constructs model, so I can mix and match:
obsMod <- 'obs =~ Q5 + Q9  + Q13 + Q17 + Q25 + Q29 + Q30 + Q33 + Q37 + Q39'
desMod <- 'des =~ Q6 + Q10 + Q14 + Q18 + Q22 + Q26 + Q34'
accMod <- 'acc =~ Q8 + Q12 + Q16 + Q20 + Q24 + Q28 + Q32 + Q36'
actMod <- 'act =~ Q7 + Q11 + Q15 + Q19 + Q23 + Q27 + Q31 + Q35 + Q38'

## Paste the construct-specific strings into a single CFA syntax string:
cfaMod <- paste(obsMod, desMod, accMod, actMod, sep = "\n")
```

---

## Basic DWLS Fit

```{r}
library(lavaan)
## Basic DWLS fit:
fit1.1 <- cfa(cfaMod,
              data = kims,
              std.lv = TRUE,
              ordered = TRUE,
              estimator = "DWLS")

summary(fit1.1)
```

---

## Robust DWLS Fit

Finally, we'll do things the right way by estimating the model with DWLS and using
robust SEs and scaled fit statistics.

```{r}
## Robust DWLS fit:
fit1.2 <- cfa(cfaMod,
              data = kims,
              std.lv = TRUE,
              ordered = TRUE,
              estimator = "WLSMV")

summary(fit1.2)
```

---

Notice that the parameter estimates are identical in the normal and robust versions
of the DWLS fit since we're using the same estimator to fit both models.

```{r}
all.equal(coef(fit1.1), coef(fit1.2))
```

---

The differences, of course, come into play with the standard errors and fit 
statistics.

```{r}
## Basic DWLS estimates:
partSummary(fit1.1, c(7:8))

## Robust DWLS estimates:
partSummary(fit1.2, c(7:8))
```

The robust SEs tend to be somewhat larger than their naive counterparts, 
especially for the latent covariance.

```{r}
## Define a few character vectors to specify which fit indices we want:
s1 <- c("chisq", "df", "pvalue")
s2 <- c("cfi", 
        "tli", 
        "rmsea", 
        "rmsea.ci.lower", 
        "rmsea.ci.upper", 
        "srmr")
s3 <- c(paste(s1, "scaled", sep = "."),
        paste(s2, "robust", sep = ".")
        )

## Fit statistics for the basic DWLS fit:
fitMeasures(fit1.1, c(s1, s2))

## Fit measures for the robust DWLS fit:
fitMeasures(fit1.2, s3)
```

Likewise, the scaled/robust fit statistics/indices suggest slightly worse fit 
than the naive versions. 

---

## Alternative `cfa()` Specifications

When estimating the last model, we actually used a bit of a shortcut. By 
setting `ordered = TRUE`, we tell the `cfa()` function that all of our observed
data are ordinal (the same will hold for `lavaan()` or any of its other wrappers). 
This approach won't work when our model also contains continuous variables. 
Fortunately, there are two alternative means of flagging ordinal variables.

---

The `ordered` argument accepts a character vector containing the names of the 
ordinal variables. 

```{r}
cfaMod <- paste(obsMod, desMod, sep = "\n")

## Explicitly name the ordinal variables:
fit1.2.1 <- cfa(cfaMod,
                data = outlook,
                std.lv = TRUE,
                ordered = paste0("Q", c(6, 10, 14, 18, 22, 26, 34)),
                estimator = "WLSMV")

summary(fit1.2.1)
```

---

# SEM

---

```{r}
semMod <- paste(cfaMod, "act ~ obs + des + acc", sep = "\n")

semFit <- sem(semMod, data = kims, ordered = TRUE, estimator = "WLSMV", std.lv = TRUE)
summary(semFit)
fitMeasures(semFit)
```

---

```{r}
semMod <- paste(obsMod, accMod, "acc ~ obs + age + male", sep = "\n")

semFit <- mutate(kims, male = ifelse(sex == "male", 1, 0)) %>% 
  sem(semMod, data = ., ordered = TRUE, estimator = "WLSMV", std.lv = TRUE)
summary(semFit)
fitMeasures(semFit)
```

---

# Measurement Invariance

---

## Configural Invariance

First, we estimate the configural invariance model by adding the grouping factor
without placing any cross-group constraints (i.e., in the same way that we did 
for continuous indicator models). 

```{r}
## Estimate the baseline (configurally invariant) model:
baseOut <- cfa(model            = cfaMod,
               data             = kims,
               group            = "sex",
               ordered          = TRUE,
               estimator        = "WLSMV",
               parameterization = "delta",
               std.lv           = TRUE,
               test             = "satorra.bentler")
```

Check the summary to ensure that everything looks sensible.

```{r}
summary(baseOut)
```

Check the model fit to test configural invariance.

```{r}
s4 <- paste(c(s1, s2), "scaled", sep = ".")

## Check the model fit:
fitMeasures(baseOut)
fitMeasures(baseOut, s4)
```

The model fits well. Configural invariance holds.

---

## Threshold Invariance

Now, we need to test threshold invariance. We do so via the same procedure we 
would use to test invariance of factor loadings or intercepts.

1. Estimate a restricted model wherein the thresholds are constrained to 
equality across groups.
1. Test the tenability of the constraint by comparing to the configurally 
invariant model.

We must establish threshold invariance before considering weak or strong invariance 
because weak and strong invariance both concern paths linking the LRVs to the 
common latent factors. We must first demonstrate, therefore, that the LRVs are 
defined equivalently in all groups.

---

We can make our lives much easier by using the `measEq.syntax()` function from 
the **semTools** package to programmatically generate the **lavaan** model syntax
for our various restricted models.

```{r}
library(semTools)

## Define the model syntax for the threshold-invariant model:
tMod <- measEq.syntax(configural.model = cfaMod,
                      data             = kims,
                      ordered          = TRUE,
                      parameterization = "delta",
                      ID.fac           = "std.lv",
                      ID.cat           = "Wu.Estabrook.2016",
                      group            = "sex",
                      group.equal      = "thresholds")
```

We can view the resulting syntax by casting the `tMod` object as a character vector
and printing it with the `cat()` function.

```{r}
tMod %>% as.character() %>% cat()
```

---

Now, we estimate the threshold invariant model. 

*Notes:*

- We do not need to specify anything for the `group.equal` argument because all 
invariance constraints are defined in the model syntax we generated via 
`measEq.syntax()`.
- We must type cast the object returned by `measEq.syntax()` as a character vector 
(i.e., by wrapping it with `as.character()`) to get the `cfa()` function to 
recognize this object as model syntax.

```{r}
## Estimate the threshold-invariant model:
tOut <- cfa(model     = as.character(tMod),
            data      = kims,
            group     = "sex",
            ordered   = TRUE,
            estimator = "WLSMV",
            test      = "satorra.bentler")

## Summarize the results:
summary(tOut)
```

As expected, the thresholds are equal across groups. Additionally, constraining 
the thresholds helps identify the model, so we can freely estimate the item 
intercepts and the scales of the latent response variables in the second and 
third groups.

```{r}
## Check the model fit:
fitMeasures(tOut, s4)
```

The model still fits very well.

---

We test threshold invariance by comparing to the configurally invariant model.

- The `semTools::compareFit()` function provides several different flavors of
model comparison.
   - Nested model $\Delta \chi^2$ test (i.e., likelihood ratio test).
   - Differences in CFI, TLI, RMSEA, and SRMR
- We specify `method = "satorra.bentler.2010"` via the `argsLRT` argument to get
the most refined $\Delta \chi^2$ test.
   - The naive difference between two scaled $\chi^2$ statistics is not $\chi^2$ 
   distributed.
   - The $\Delta \chi^2$ test needs to be conducted with particular corrections
   to account for the scaled fit statistics.
   
```{r}
## Test threshold invariance via model comparisons:
compareFit(baseOut, 
           tOut, 
           argsLRT = list(method = "satorra.bentler.2010")
           ) %>% 
  summary()
```

The $\Delta \chi^2$ test is significant, so we should reject the null hypothesis 
of invariant thresholds. We have not established threshold invariance.

- The $\Delta \mathrm{CFI}$ and $\Delta \mathrm{RMSEA}$ are both quite small, but
the methodological work evaluating measurement invariance testing with approximate 
fit indices has not considered the robust/scaled versions of these indices. So, 
we don't have much basis for using these indices for measurement invariance testing. 

---

Though we're not really justified in doing so, we'll continue with the process of
measurement invariance testing, for didactic purposes.

---

## Weak Invariance

After establishing threshold invariance, the next step is testing the additional 
constraint of loading invariance (i.e., weak invariance or metric invariance).

- Again, we can use the `measEq.syntax()` function to streamline the process.

```{r}
## Define the model syntax for the threshold/loading-invariant model:
tlMod <- measEq.syntax(configural.model = cfaMod,
                       data             = outlook,
                       ordered          = TRUE,
                       parameterization = "delta",
                       ID.fac           = "std.lv",
                       ID.cat           = "Wu.Estabrook.2016",
                       group            = "party",
                       group.equal      = c("thresholds", "loadings")
                       )

## Check the resulting syntax:
tlMod %>% as.character() %>% cat()
```

---

We estimate the model in the usual way.

```{r}
## Estimate the threshold/loading-invariant model:
tlOut <- cfa(model     = as.character(tlMod),
             data      = outlook,
             group     = "party",
             ordered   = TRUE,
             estimator = "WLSMV")

## Summarize the results:
summary(tlOut)
```

Now that we've equated the factor loadings across groups, we have the additional
identification necessary to freely estimate the latent variances in the second 
and third groups.

```{r}
## Check the model fit:
fitMeasures(tlOut, s4)
```

Again, model fit is still excellent.

---

We evaluate weak invariance by comparing back to the threshold-invariance model.

```{r}
## Test threshold/loading invariance via model comparisons:
compareFit(tOut, 
           tlOut, 
           argsLRT = list(method = "satorra.bentler.2010")
           ) %>% 
  summary()
```

Relative to the threshold-invariant model, we haven't lost a significant amount
of fit by equating the factor loadings across groups. Conditioning on the dubious
threshold invariance, we can claim weak invariance.

---

## Strong Invariance

We conclude by equating the item intercepts across groups to test strong/metric
invariance.

```{r}
## Define the model syntax for the threshold/loading/intercept-invariant model:
tliMod <-
    measEq.syntax(configural.model = cfaMod,
                  data             = outlook,
                  ordered          = TRUE,
                  parameterization = "delta",
                  ID.fac           = "std.lv",
                  ID.cat           = "Wu.Estabrook.2016",
                  group            = "party",
                  group.equal      = c("thresholds", "loadings", "intercepts")
                  )

## Check the syntax:
tliMod %>% as.character() %>% cat()

## Estimate the threshold/loading/intercept-invariant model:
tliOut <- cfa(model     = as.character(tliMod),
              data      = outlook,
              group     = "party",
              ordered   = TRUE,
              estimator = "WLSMV")

## Summarize the results:
summary(tliOut)
```

Constraining the item intercepts allows us to freely estimate the latent means
in the second and third groups.

```{r}
## Check the model fit:
fitMeasures(tliOut, s4)
```

The strongly invariant model still fits very well.

```{r}
## Test threshold/loading/intercept invariance via model comparisons:
compareFit(tlOut, 
           tliOut, 
           argsLRT = list(method = "satorra.bentler.2010")
           ) %>% 
  summary()
```

We don't lose a significant amount of fit by constraining the item intercepts in
addition to the factor loadings and thresholds. Given our acceptance of the weakly 
invariant model, we can claim strong invariance.

---

# Concluding Remarks

---

The only substantial differences between modeling ordinal indicators via CVM and
modeling continuous indicators via normal-theory ML concern the measurement model. 
So, after adapting the CFA as described above, all the structural modeling and 
testing works pretty much as it does for ordinary normal-theory models.

There are, however, two frequently applicable caveats to the above statement.

1. As you've seen above, the difference in two scaled $\chi^2$ statistics is not 
$\chi^2$ distributed. So, you must apply an appropriate correction to any 
$\Delta \chi^2$ tests. 
   - Fortunately, these corrections are applied automatically when you conduct 
   the tests with `anova()`, `lavTestLRT()`, or `compareFit()`.
1. The WLS and DWLS estimators cannot handle missing data (i.e., there is no WLS 
analog to FIML). So, you must treat any missing prior to model estimation. With 
FIML off the table, you should use multiple imputation (MI) unless you have a
trivial amount of missing data (in which case a single stochastic regression 
imputation will suffice).
   - Again, **semTools** comes to the rescue here. The `lavaan.mi()` routines 
   covered in Module 6 make the process of conducting the above analyses with 
   multiply imputed data relatively painless.

---

End of Demonstration

[outlook0]: https://doi.org/10.3886/ICPSR35348.v1
[outlook1]: https://github.com/kylelang/lavaan-e-learning/raw/main/data/outlook.rds
