---
title: "Lab 8: Categorical Indicators" 
subtitle: "SEM in R using lavaan"
author: "Kyle M. Lang"
date: "Updated: `r format(Sys.time(), format = '%Y-%m-%d')`"
params:
  answers: true
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 1
    toc_float: true
    number_sections: true
    df_print: paged
    css: "../../resources/style.css"
editor_options: 
  chunk_output_type: console
bibliography: ../../bibtex/data_refs.bib
csl: ../../bibtex/apa.csl
link-citations: yes
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(dplyr)
library(magrittr)
library(ggplot2)

source(here::here("code", "supportFunctions.R"))

figDir <- "../figures/"

set.seed(235711)

## Define an asis engine that will evaluate inline code within an asis block:
knit_engines$set(asis = function(options) {
  if(options$echo && options$eval) knit_child(text = options$code)
}
)

opts_chunk$set(
  include = params$answers, 
  echo = params$answer, 
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  comment = NA,
  root.dir = here::here("./")
)
```

<!-- 
Define some hacky LaTeX commands to force nice spacing between lines
NOTE: These must be called within a math environment (e.g., $\va$)
-->
\newcommand{\va}{\\[12pt]}
\newcommand{\vb}{\\[6pt]}
\newcommand{\vc}{\\[3pt]}
\newcommand{\vx}[1]{\\[#1pt]}

---

In this lab, you'll explore methods of doing CFA/SEM with categorical data using **lavaan**.

---

# Data

---

We'll be analyzing data collected with the *Kentucky Inventory of Mindfulness Skills* [@baerEtAl:2004]. You can
download the dataset that we'll analyze below via [this link][kims1], and you can access the original data
[here][kims1]. The KIMS contains 39 items assessing four dimensions of mindfulness skills: *Observing*, *Describing*,
*Acting*, and *Accepting*. The following list shows the putative construct-to-item mapping.

**Observing**

- `Q1`: I notice changes in my body, such as whether my breathing slows down or speeds up. 
- `Q5`: I pay attention to whether my muscles are tense or relaxed. 
- `Q9`: When I'm walking, I deliberately notice the sensations of my body moving. 
- `Q13`: When I take a shower or bath, I stay alert to the sensations of water on my body. 
- `Q17`: I notice how foods and drinks affect my thoughts, bodily sensations, and emotions. 
- `Q25`: I pay attention to sounds, such as clocks ticking, birds chirping, or cars passing. 
- `Q29`: I notice the smells and aromas of things. 
- `Q30`: I intentionally stay aware of my feelings. 
- `Q33`: I notice visual elements in art or nature, such as colors, shapes, textures, or patterns of light and shadow. 
- `Q37`: I pay attention to how my emotions affect my thoughts and behavior. 
- `Q39`: I notice when my moods begin to change.

**Describing**

- `Q2`: I'm good at finding the words to describe my feelings. 
- `Q6`: I can easily put my beliefs, opinions, and expectations into words. 
- `Q10`: I'm good at thinking of words to express my perceptions, such as how things taste, smell, or sound. 
- `Q14`: It's hard for me to find the words to describe what I'm thinking. 
- `Q18`: I have trouble thinking of the right words to express how I feel about things. 
- `Q22`: When I have a sensation in my body, it's difficult for me to describe it because I cant find the right words. 
- `Q26`: Even when I'm feeling terribly upset, I can find a way to put it into words. 
- `Q34`: My natural tendency is to put my experiences into words. 

**Acting**

- `Q3`: When I do things, my mind wanders off and I'm easily distracted. 
- `Q7`: When I'm doing something, I'm only focused on what I'm doing, nothing else. 
- `Q11`: I drive on "automatic pilot" without paying attention to what I'm doing. 
- `Q15`: When I'm reading, I focus all my attention on what I'm reading. 
- `Q19`: When I do things, I get totally wrapped up in them and don't think about anything else. 
- `Q23`: I don't pay attention to what I'm doing because I'm daydreaming, worrying, or otherwise distracted. 
- `Q27`: When I'm doing chores, such as cleaning or laundry, I tend to daydream or think of other things. 
- `Q31`: I tend to do several things at once rather than focusing on one thing at a time. 
- `Q35`: When I'm working on something, part of my mind is occupied with other topics, such as what I'll be doing later, or things I'd rather be doing. 
- `Q38`: I get completely absorbed in what I'm doing, so that all my attention is focused on it. 

**Accepting**

- `Q4`: I criticize myself for having irrational or inappropriate emotions. 
- `Q8`: I tend to evaluate whether my perceptions are right or wrong. 
- `Q12`: I tell myself that I shouldn't be feeling the way I'm feeling. 
- `Q16`: I believe some of my thoughts are abnormal or bad and I shouldn't think that way.  
- `Q20`: I make judgments about whether my thoughts are good or bad. 
- `Q24`: I tend to make judgments about how worthwhile or worthless my experiences are. 
- `Q28`: I tell myself that I shouldn't be thinking the way I'm thinking. 
- `Q32`: I think some of my emotions are bad or inappropriate and I shouldn't feel them. 
- `Q36`: I disapprove of myself when I have irrational ideas. 

These data were collected via a self-report questionnaire wherein all the KIMS items were rated on a five-point
Likert-type scale. So, we'll be better off using CVM-based modeling to analyzing these data as ordinal, rather than
pretending that the items are normally distributed.

---

##

**Load the *kims.rds* dataset.**

```{r}
dataDir <- "data"
kims    <- readRDS(here::here(dataDir, "kims.rds"))
```

---

# Categorical Variable CFA

---

Let's assume we want to use the KIMS data to estimate a three-factor CFA model of the *Observing*, *Describing*, and
*Accepting* constructs. For these exercises, we'll use the item numbers listed below to define the three latent
constructs in our CFA.

- **Observing:** 5, 9, 17, 25, 29, 30, 37
- **Describing:** 2, 6, 10, 14, 18, 22, 26, 34
- **Accepting:** 4, 12, 20, 24, 28, 32, 36

---

##

**Define the model syntax for the three-factor CFA described above.**

```{r}
cfaMod <- '
obs =~ Q5 + Q9  + Q17 + Q25 + Q29 + Q30 + Q37
des =~ Q2 + Q6  + Q10 + Q14 + Q18 + Q22 + Q26 + Q34
acc =~ Q4 + Q12 + Q20 + Q24 + Q28 + Q32 + Q36
'
```

---

##

**Estimate the CFA using appropriate CVM-based settings.**

- Allow all latent constructs to covary.
- Set the scale with the fixed-factor method.
- Estimate the model with DWLS.
- Use robust SEs to quantify sampling variability.
- Use the scaled-and-shifted $\chi^2$ statistic to evaluate model fit.

```{r}
library(lavaan)

## Specify individual options:
cfaOut <- cfa(cfaMod,
              data = kims,
              std.lv = TRUE,
              ordered = TRUE,
              estimator = "DWLS",
              se = "robust",
              test = "scaled.shifted")

## OR ##

## Use the WLSMV shorthand:
cfaOut2 <- cfa(cfaMod,
              data = kims,
              std.lv = TRUE,
              ordered = TRUE,
              estimator = "WLSMV")
```

---

## {#cfa}

**Summarize and evaluate the model.**

- Is the model an acceptable representation of the data?
   - Does the model fit well?
   - Do the parameter estimates seem plausible?

```{r}
## Summarize the model:
summary(cfaOut)

## Define a character to specify only the scaled fit indices:
stats <- paste(
  c("chisq",
    "df",
    "pvalue",
    "cfi", 
    "tli", 
    "rmsea", 
    "rmsea.ci.lower", 
    "rmsea.ci.upper"),
  "scaled",
  sep = "."
)

## Fit measures for the robust DWLS fit:
fitMeasures(cfaOut, stats)
```

```{asis}
Yes, the model fits the data well and the parameter estimates all appear plausible. This CFA seems like a reasonable
representation of the measurement structure underlying these data.
```

---

# SEM

---

##

**Estimate a baseline SEM wherein *Accepting* is regressed onto *Age*.**

- Estimate the covariances between *Age*, *Describing*, and *Observing*.
- Do not allow *Describing* or *Observing* to covary with *Accepting*.

```{r}
## Add the structural paths to the CFA syntax:
semMod <- paste(
  cfaMod,
  "acc ~ age",
  "age ~~ des + obs",
  sep = "\n"
)

## Tell lavaan that all variable other than 'age' are ordinal:
cats <- setdiff(colnames(kims), "age")

## Estimate the model:
semFit0 <- sem(
  semMod,
  data = kims,
  std.lv = TRUE,
  ordered = cats,
  estimator = "WLSMV"
)
```

---

##

**Summarize and interpret the model.**

- Does *Age* significantly predict *Accepting*?
- If so, interpret the age effect.

```{r}
summary(semFit0)
```

```{r, include = FALSE}
b <- inspect(semFit0, "est")$beta["acc", "age"]
s <- inspect(semFit0, "se")$beta["acc", "age"]
z <- b / s
p <- pnorm(abs(z), lower.tail = FALSE) * 2
```

```{asis}
Yes, *Age* significantly predicts *Accepting*
$(\beta = `r round(b, 3)`,$
$\mathit{SE} = `r round(s, 3)`,$
$z = `r round(z, 3)`,$
$`r prettyPValue(p)`).$
Older tend to have significantly higher levels of accepting than younger participants.
```

---

##

**Estimate an augmented version of the model from above.**

- Add *Observing* and *Describing* as predictors of *Accepting*.
- Do not change anything else about the model structure.

```{r}
semMod <- paste(
  cfaMod,
  "acc ~ age + des + obs",
  "age ~~ des + obs",
  sep = "\n"
)

semFit1 <- sem(
  semMod,
  data = kims,
  std.lv = TRUE,
  ordered = cats,
  estimator = "WLSMV"
)
```

---

##

**Summarize and interpret the model parameters.**

- Does adding the two new predictors significantly improve the model?
- Interpret the effects of *Observing* and *Describing*.

```{r}
summary(semFit1)
```

```{r, paged.print = FALSE}
anova(semFit0, semFit1)
```

```{r, include = FALSE}
test <- anova(semFit0, semFit1)

x2 <- test[2, "Chisq diff"] |> round(2)
df <- test[2, "Df diff"]
px <- prettyPValue(test[2, "Pr(>Chisq)"])

b <- inspect(semFit1, "est")$beta["acc", c("obs", "des")]
s <- inspect(semFit1, "se")$beta["acc", c("obs", "des")]
z <- b / s
pb <- pnorm(abs(z), lower.tail = FALSE) * 2
```

```{asis}
Yes, adding *Observing* and *Describing* significantly improves the model
$(\Delta \chi^2[`r df`] = `r x2`,$
$`r px`).$
After controlling for *Age* and *Observing*, *Describing* has a significant, negative effect on *Accepting*
$(\beta = `r round(b[["des"]], 3)`,$
$\mathit{SE} = `r round(s[["des"]], 3)`,$
$z = `r round(z[["des"]], 3)`,$
$`r prettyPValue(pb[["des"]])`).$
*Observing*, on the other hand, is not a signifiant predictor of *Accepting*, after controlling for *Age* and
*Describing*
$(\beta = `r round(b[["obs"]], 3)`,$
$\mathit{SE} = `r round(s[["obs"]], 3)`,$
$z = `r round(z[["obs"]], 3)`,$
$`r prettyPValue(pb[["obs"]])`).$
```

---

# Measurement Invariance

---

##

**Estimate the CFA from \@ref(cfa) as a multiple-group model.**

- Specify *sex* as the grouping factor.
- Estimate a configurally invariant model (i.e., don't equate any parameters across groups).

```{r}
## Estimate the baseline (configurally invariant) model:
baseOut <- cfa(model            = cfaMod,
               data             = kims,
               std.lv           = TRUE,
               group            = "sex",
               ordered          = TRUE,
               estimator        = "WLSMV",
               parameterization = "delta")

## Check the summary to ensure that everything looks sensible.
summary(baseOut)
```

---

## 

**Evaluate configural invariance.**

- What is your conclusion?

```{r}
## Check the model fit:
fitMeasures(baseOut, stats)
```

```{asis}
The model fits well. Configural invariance holds.
```

---

##

**Define the model syntax for the threshold invariance model.**

- Use the `measEq.syntax()` function from the **semTools** package to generate the model syntax.
- Set the scale by standardizing the latent factors (i.e., use the fixed-factor/reference-group method of identification).
- Use the delta parameterization to define the latent response variables.
- Use the Wu & Estabrook (2016) identification conditions.

```{r}
library(semTools)

## Define the model syntax for the threshold-invariant model:
tMod <- measEq.syntax(configural.model = cfaMod,
                      data             = kims,
                      ordered          = TRUE,
                      parameterization = "delta",
                      ID.fac           = "std.lv",
                      ID.cat           = "Wu.Estabrook.2016",
                      group            = "sex",
                      group.equal      = "thresholds")

## Print the syntax to check:
tMod |> as.character() |> cat()
```

---

##

**Estimate the threshold invariant model.**

- Check the results by summarizing the model and evaluating the model fit.

```{r}
## Estimate the threshold-invariant model:
tOut <- cfa(model     = as.character(tMod),
            data      = kims,
            group     = "sex",
            ordered   = TRUE,
            estimator = "WLSMV")

## Summarize the results:
summary(tOut)

## Check the model fit:
fitMeasures(tOut, stats)
```

```{asis}
Everything looks good. The thresholds are correctly equated across groups. All the parameter estimates seem sensible,
and the model still fits well.
```

---

##

**Test threshold invariance.**

```{r}
## Test threshold invariance via model comparisons:
compareFit(baseOut, tOut) |> summary()
```

```{r, include = FALSE}
tmp  <- compareFit(baseOut, tOut)
test <- tmp@nested

x2 <- test[2, "Chisq diff"] |> round(2)
df <- test[2, "Df diff"]
p  <- prettyPValue(test[2, "Pr(>Chisq)"])
```

```{asis}
The $\Delta \chi^2$ test is nonsignificant
$(\Delta \chi^2[`r df`] = `r x2`,$
$`r p`),$
so we can infer invariant thresholds.
```

---

##

**Define the model syntax for the weak/metric invariance model.**

- Use the `measEq.syntax()` function from the **semTools** package to generate the model syntax.
- Set the scale by standardizing the latent factors (i.e., use the fixed-factor/reference-group method of identification).
- Use the delta parameterization to define the latent response variables.
- Use the Wu & Estabrook (2016) identification conditions.

```{r}
## Define the model syntax for the threshold/loading-invariant model:
tlMod <- measEq.syntax(configural.model = cfaMod,
                       data             = kims,
                       ordered          = TRUE,
                       parameterization = "delta",
                       ID.fac           = "std.lv",
                       ID.cat           = "Wu.Estabrook.2016",
                       group            = "sex",
                       group.equal      = c("thresholds", "loadings")
                       )

## Check the resulting syntax:
tlMod |> as.character() |> cat()
```

---

##

**Estimate the weak invariant model.**

- Check the results by summarizing the model and evaluating the model fit.

```{r}
## Estimate the threshold/loading-invariant model:
tlOut <- cfa(model     = as.character(tlMod),
             data      = kims,
             group     = "sex",
             ordered   = TRUE,
             estimator = "WLSMV")

## Summarize the results:
summary(tlOut)

## Check the model fit:
fitMeasures(tlOut, stats)
```

```{asis}
Again, model fit is still good, and all the parameter estimates are reasonable.
```

---

## {#weakInvariance}

**Test weak invariance.**

```{r}
## Test threshold/loading invariance via model comparisons:
compareFit(tOut, tlOut) |> summary()
```

```{r, include = FALSE}
tmp  <- compareFit(tOut, tlOut)
test <- tmp@nested

x2 <- test[2, "Chisq diff"] |> round(2)
df <- test[2, "Df diff"]
p  <- prettyPValue(test[2, "Pr(>Chisq)"])
```

```{asis}
Again, the $\Delta \chi^2$ test is nonsignificant
$(\Delta \chi^2[`r df`] = `r x2`,$
$`r p`),$
so we can infer weak invariance.
```

---

##

**Define the model syntax for the strong/scalar invariance model.**

- Use the `measEq.syntax()` function from the **semTools** package to generate the model syntax.
- Set the scale by standardizing the latent factors (i.e., use the fixed-factor/reference-group method of identification).
- Use the delta parameterization to define the latent response variables.
- Use the Wu & Estabrook (2016) identification conditions.

```{r}
## Define the model syntax for the threshold/loading/intercept-invariant model:
tliMod <-
    measEq.syntax(configural.model = cfaMod,
                  data             = kims,
                  ordered          = TRUE,
                  parameterization = "delta",
                  ID.fac           = "std.lv",
                  ID.cat           = "Wu.Estabrook.2016",
                  group            = "sex",
                  group.equal      = c("thresholds", "loadings", "intercepts")
                  )

## Check the syntax:
tliMod |> as.character() |> cat()
```

---

##

**Estimate the strong invariance model.**

- Check the results by summarizing the model and evaluating the model fit.

```{r}
## Estimate the threshold/loading/intercept-invariant model:
tliOut <- cfa(model     = as.character(tliMod),
              data      = kims,
              group     = "sex",
              ordered   = TRUE,
              estimator = "WLSMV")

## Summarize the results:
summary(tliOut)

## Check the model fit:
fitMeasures(tliOut, stats)
```

```{asis}
The strongly invariant model still fits well, and the parameter estimates remain sensible.
```

---

##

**Test strong invariance.**

```{r}
## Test threshold/loading/intercept invariance via model comparisons:
compareFit(tlOut, tliOut) |> summary()
```

```{r, include = FALSE}
tmp  <- compareFit(tlOut, tliOut)
test <- tmp@nested

x2 <- test[2, "Chisq diff"] |> round(2)
df <- test[2, "Df diff"]
p  <- prettyPValue(test[2, "Pr(>Chisq)"])
```

```{asis}
This time, the $\Delta \chi^2$ test is significant
$(\Delta \chi^2[`r df`] = `r x2`,$
$`r p`),$
so we must reject the null hypothesis of strongly invariant measurement. Constraining the item intercepts, in addition
to the factor loadings and thresholds, produces an unacceptable loss of model fit.
```

---

# Moderated SEM

---

In this section, you will use a multiple-group SEM to test the following moderation hypotheses:

- $H_1$: The partial effect of *Describing* on *Accepting* (controlling for *Age* and *Observing*) differs for males
and females.
- $H_2$: The partial effect of *Observing* on *Accepting* (controlling for *Age* and *Describing*) differs for males
and females.

---

##

**Estimate the unrestricted, moderated SEM.**

- Label the potentially moderated parameters.
- Maintain the weak invariance constraints imposed in \@ref(weakInvariance).

```{r}
## Define model syntax with labels for the potentially moderated effects:
semMod <- paste(
  cfaMod,
  'acc ~ age + c("b2f", "b2m") * des + c("b3f", "b3m") * obs',
  'age ~~ des + obs',
  sep = "\n"
)

## Estimate the unrestricted moderated model:
modOut <- sem(
  model = semMod,
  data = kims,
  group = "sex",
  ordered = cats,
  estimator = "WLSMV",
  parameterization = "delta",
  std.lv = TRUE,
  group.equal = c("thresholds", "loadings")
)

summary(modOut)
```

---

## {#h1Test}

**Use `lavaan::lavTestWald()` to test $H_1$.**

- What is your conclusion?

```{r}
lavTestWald(modOut, constraints = "b2f == b2m")
```

```{r, include = FALSE}
test1 <- lavTestWald(modOut, constraints = "b2f == b2m")
```

```{asis}
$H_1$ is not supported
$(\Delta \chi^2[`r test1$df`] = `r round(test1$stat, 3)`,$
$`r prettyPValue(test1$p.value)`).$
There is not a significant sex difference in the partial effect of *Describing* on *Accepting*.
```

---

## {#h2Test}

**Use `lavaan::lavTestWald()` to test $H_2$.**

- What is your conclusion?

```{r}
lavTestWald(modOut, constraints = "b3f == b3m")
```

```{r, include = FALSE}
test2 <- lavTestWald(modOut, constraints = "b3f == b3m")
```

```{asis}
$H_2$ is supported
$(\Delta \chi^2[`r test2$df`] = `r round(test2$stat, 3)`,$
$`r prettyPValue(test2$p.value)`).$
*Sex* significantly affects the partial effect of *Observing* on *Accepting*.
```

---

## {#parsimoniousSem}

**Estimate a parsimonious model.**

- Apply any parameter constraints suggested by the test results from \@ref(h1Test) and \@ref(h2Test).

```{r}
## Equate the 'describing' effect across groups:
semMod <- paste(
  cfaMod,
  'acc ~ age + c("b2", "b2") * des + c("b3f", "b3m") * obs',
  'age ~~ des + obs',
  sep = "\n"
)

## Estimate the parsimonious model:
modOut2 <- sem(
  model = semMod,
  data = kims,
  group = "sex",
  ordered = cats,
  estimator = "WLSMV",
  parameterization = "delta",
  std.lv = TRUE,
  group.equal = c("thresholds", "loadings")
)
```

```{asis}
The hypothesis tests we conducted above suggested no sex difference for the *Describing* effect, so we can simplify our
model by equating the partial effect of *Describing* on *Accepting* across groups. We can double-check the tenability of
this constraint with a nested model comparison test.
```

```{r, paged.print = FALSE}
anova(modOut, modOut2)
```

```{asis}
As expected, the $\Delta \chi^2$ statistic is nonsignificant. Note how the p-value from this test is very similar to the
p-value from the test of $H_1$ that we conducted in \@ref(h1Test). Both of these tests are evaluating the same
hypothesis, so we shouldn't be surprised by the equivalent results.
```

---

##

**Interpret the final *Observing* and *Describing* effects.**

- Interpret the estimates from the parsimonious model you estimated in \@ref(parsimoniousSem).

```{r}
## Summarize the parsimonous model:
summary(modOut2)
```

```{r, include = FALSE}
bf <- inspect(modOut2, "est")$female$beta["acc", c("obs", "des")]
bm <- inspect(modOut2, "est")$male$beta["acc", c("obs", "des")]
sf <- inspect(modOut2, "se")$female$beta["acc", c("obs", "des")]
sm <- inspect(modOut2, "se")$male$beta["acc", c("obs", "des")]

zf <- bf / sf
zm <- bm / sm

pf <- pnorm(abs(zf), lower.tail = FALSE) * 2
pm <- pnorm(abs(zm), lower.tail = FALSE) * 2
```

```{asis}
After controlling for *Age* and *Observing*, *Describing* has a significant, negative effect on *Accepting*
$(\beta = `r round(bf[["des"]], 3)`,$
$\mathit{SE} = `r round(sf[["des"]], 3)`,$
$z = `r round(zf[["des"]], 3)`,$
$`r prettyPValue(pf[["des"]])`),$
and this effect does not significantly differ for males and females
$(\Delta \chi^2[`r test1$df`] = `r round(test1$stat, 3)`,$
$`r prettyPValue(test1$p.value)`).$

On the other hand, $Sex$ has significant impact on the partial *Observing* effect
$(\Delta \chi^2[`r test2$df`] = `r round(test2$stat, 3)`,$
$`r prettyPValue(test2$p.value)`).$
After controlling for *Age* and *Describing*, *Observing* has a significant, positive effect on *Accepting* for males
$(\beta = `r round(bm[["obs"]], 3)`,$
$\mathit{SE} = `r round(sm[["obs"]], 3)`,$
$z = `r round(zm[["obs"]], 3)`,$
$`r prettyPValue(pm[["obs"]])`),$
but this effect is nonsignificant for females
$(\beta = `r round(bf[["obs"]], 3)`,$
$\mathit{SE} = `r round(sf[["obs"]], 3)`,$
$z = `r round(zf[["obs"]], 3)`,$
$`r prettyPValue(pf[["obs"]])`).$
```

---

End of Lab 8

---

[kims0]: https://www.kaggle.com/datasets/lucasgreenwell/kentucky-inventory-of-mindfulness-skills-responses
[kims1]: https://github.com/kylelang/lavaan-e-learning/raw/main/data/kims.rds
