---
title: "Lab 8: Categorical Indicators" 
subtitle: "SEM in R using lavaan"
author: "Kyle M. Lang"
date: "Updated: `r format(Sys.time(), format = '%Y-%m-%d')`"
params:
  answers: true
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    df_print: paged
    css: "../../resources/style.css"
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(dplyr)
library(magrittr)
library(ggplot2)

source(here::here("code", "supportFunctions.R"))

figDir <- "../figures/"

set.seed(235711)

## Define an asis engine that will evaluate inline code within an asis block:
knit_engines$set(asis = function(options) {
  if(options$echo && options$eval) knit_child(text = options$code)
}
)

opts_chunk$set(
  include = params$answers, 
  echo = params$answer, 
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  comment = NA,
  root.dir = here::here("./")
)
```

<!-- 
Define some hacky LaTeX commands to force nice spacing between lines
NOTE: These must be called within a math environment (e.g., $\va$)
-->
\newcommand{\va}{\\[12pt]}
\newcommand{\vb}{\\[6pt]}
\newcommand{\vc}{\\[3pt]}
\newcommand{\vx}[1]{\\[#1pt]}

---

In this lab, you'll explore methods of doing CFA/SEM with categorical data using **lavaan**.

---

# Data

---

**Observing**

- `Q1`: I notice changes in my body, such as whether my breathing slows down or speeds up. 
- `Q5`: I pay attention to whether my muscles are tense or relaxed. 
- `Q9`: When I'm walking, I deliberately notice the sensations of my body moving. 
- `Q13`: When I take a shower or bath, I stay alert to the sensations of water on my body. 
- `Q17`: I notice how foods and drinks affect my thoughts, bodily sensations, and emotions. 
- `Q25`: I pay attention to sounds, such as clocks ticking, birds chirping, or cars passing. 
- `Q29`: I notice the smells and aromas of things. 
- `Q30`: I intentionally stay aware of my feelings. 
- `Q33`: I notice visual elements in art or nature, such as colors, shapes, textures, or patterns of light and shadow. 
- `Q37`: I pay attention to how my emotions affect my thoughts and behavior. 
- `Q39`: I notice when my moods begin to change.

**Describing**

- `Q2`: I'm good at finding the words to describe my feelings. 
- `Q6`: I can easily put my beliefs, opinions, and expectations into words. 
- `Q10`: I'm good at thinking of words to express my perceptions, such as how things taste, smell, or sound. 
- `Q14`: It's hard for me to find the words to describe what I'm thinking. 
- `Q18`: I have trouble thinking of the right words to express how I feel about things. 
- `Q22`: When I have a sensation in my body, it's difficult for me to describe it because I cant find the right words. 
- `Q26`: Even when I'm feeling terribly upset, I can find a way to put it into words. 
- `Q34`: My natural tendency is to put my experiences into words. 

**Acting**

- `Q3`: When I do things, my mind wanders off and I'm easily distracted. 
- `Q7`: When I'm doing something, I'm only focused on what I'm doing, nothing else. 
- `Q11`: I drive on "automatic pilot" without paying attention to what I'm doing. 
- `Q15`: When I'm reading, I focus all my attention on what I'm reading. 
- `Q19`: When I do things, I get totally wrapped up in them and don't think about anything else. 
- `Q23`: I don't pay attention to what I'm doing because I'm daydreaming, worrying, or otherwise distracted. 
- `Q27`: When I'm doing chores, such as cleaning or laundry, I tend to daydream or think of other things. 
- `Q31`: I tend to do several things at once rather than focusing on one thing at a time. 
- `Q35`: When I'm working on something, part of my mind is occupied with other topics, such as what I'll be doing later, or things I'd rather be doing. 
- `Q38`: I get completely absorbed in what I'm doing, so that all my attention is focused on it. 

**Accepting**

- `Q4`: I criticize myself for having irrational or inappropriate emotions. 
- `Q8`: I tend to evaluate whether my perceptions are right or wrong. 
- `Q12`: I tell myself that I shouldn't be feeling the way I'm feeling. 
- `Q16`: I believe some of my thoughts are abnormal or bad and I shouldn't think that way.  
- `Q20`: I make judgments about whether my thoughts are good or bad. 
- `Q24`: I tend to make judgments about how worthwhile or worthless my experiences are. 
- `Q28`: I tell myself that I shouldn't be thinking the way I'm thinking. 
- `Q32`: I think some of my emotions are bad or inappropriate and I shouldn't feel them. 
- `Q36`: I disapprove of myself when I have irrational ideas. 

---

First, we'll read in the *kims.rds* dataset, and save the resulting data frame as `kims`.

```{r}
dataDir <- "data"
kims    <- readRDS(here::here(dataDir, "kims.rds"))
```

---

# Categorical Variable CFA

---

Let's assume we want to fit a three-factor CFA model of the *Observing*, *Describing*, and *Accepting* constructs
represented in the KIMS data.

##

**Define the model syntax for the three-factor CFA described above.**

```{r}
cfaMod <- '
obs =~ Q1 + Q5 + Q9  + Q13 + Q17 + Q25 + Q29 + Q30 + Q33 + Q37 + Q39
des =~ Q2 + Q6 + Q10 + Q14 + Q18 + Q22 + Q26 + Q34
acc =~ Q5 + Q8 + Q12 + Q16 + Q20 + Q24 + Q28 + Q32 + Q36
'
```

---

##

**Estimate the CFA using appropriate CVM-based settings.**

- Estimate the model with DWLS
- Use robust SEs to quantify sampling variability
- Use the scaled-and-shifted $\chi^2$ statistic to evaluate model fit

```{r}
library(lavaan)

## Specify individual options:
cfaOut <- cfa(cfaMod,
              data = kims,
              std.lv = TRUE,
              ordered = TRUE,
              estimator = "DWLS",
              se = "robust",
              test = "scaled.shifted")

## OR ##

## Use the WLSMV shorthand:
cfaOut2 <- cfa(cfaMod,
              data = kims,
              std.lv = TRUE,
              ordered = TRUE,
              estimator = "WLSMV")
```

---

##

**Summarize and evalute the model.**

- Is the model an acceptable representaiton of the data?
   - Does the model fit well?
   - Do the parameter estimates seem plausible?

```{r}
summary(cfaOut)

## Define a character to specify only the scaled fit indices:
stats <- paste(
  c("chisq",
    "df",
    "pvalue",
    "cfi", 
    "tli", 
    "rmsea", 
    "rmsea.ci.lower", 
    "rmsea.ci.upper"),
  "scaled",
  sep = "."
)

## Fit measures for the robust DWLS fit:
fitMeasures(cfaOut, stats)
```

---

# SEM

---

##

**Estimate a baseline SEM wherein *Accepting* is regressed onto *Age*.**

- Estimate the covariances between *Age*, *Describing*, and *Observing*.
- Do not allow *Describing* or *Observing* to covary with *Accepting*.

```{r}
semMod <- paste(
  cfaMod,
  "acc ~ age",
  "age ~~ des + obs",
  sep = "\n"
)

cats <- setdiff(colnames(kims), "age")

semFit0 <- sem(semMod, data = kims, ordered = cats, estimator = "WLSMV", std.lv = TRUE)
```

##

**Summarize and interpret the model.**

- Does *Age* significantly predict *Accepting*?
- If so, interpret the age effect.

```{r}
summary(semFit0)
```

---

##

**Estimate an augmented version of the model from above.**

- Add *Observing* and *Describing* as predictors of *Accepting*.
- Do not change anything else about the model structure.

```{r}
semMod <- paste(
  cfaMod,
  "acc ~ age + des + obs",
  "age ~~ des + obs",
  sep = "\n"
)

semFit1 <- sem(semMod, data = kims, ordered = cats, estimator = "WLSMV", std.lv = TRUE)
```

---

##

**Summarize and interpret the model parameters.**

- Does adding the two new predictors significantly improve the model?
- Interpret the effects of *Observing* and *Describing*.

```{r}
summary(semFit1)
fitMeasures(semFit1, stats)

anova(semFit0, semFit1)
```

---

# Measurement Invariance

---

##

**Estimate the CFA from XXX as a mutiple-group model.**

- Specify *sex* as the grouping factor.
- Estimate a configurally invariant model (i.e., don't equate any parameters across groups).

```{r}
## Estimate the baseline (configurally invariant) model:
baseOut <- cfa(model            = cfaMod,
               data             = kims,
               group            = "sex",
               ordered          = TRUE,
               estimator        = "WLSMV",
               parameterization = "delta",
               std.lv           = TRUE)

## Check the summary to ensure that everything looks sensible.
summary(baseOut)
```

---

## 

**Evaluate configural invariance.**

- What is your conclusion?

```{r}
## Check the model fit:
fitMeasures(baseOut, stats)
```

The model fits well. Configural invariance holds.

---

##

**Define the model syntax for the threshold invariance model.**

- Use the `measEq.syntax()` function from the **semTools** package to generate the model syntax.
- Set the scale by standardizing the latent factors (i.e., use the fixed-factor/reference-group method of identification).
- Use the delta parameterization to define the latent response variables.
- Use the Wu & Estabrook (2016) identification conditions.

```{r}
library(semTools)

## Define the model syntax for the threshold-invariant model:
tMod <- measEq.syntax(configural.model = cfaMod,
                      data             = kims,
                      ordered          = TRUE,
                      parameterization = "delta",
                      ID.fac           = "std.lv",
                      ID.cat           = "Wu.Estabrook.2016",
                      group            = "sex",
                      group.equal      = "thresholds")

## Print the syntax to check:
tMod |> as.character() |> cat()
```

---

##

**Estimate the threshold invariant model.**

- Check the results by summarizing the model and evaluating the model fit.

```{r}
## Estimate the threshold-invariant model:
tOut <- cfa(model     = as.character(tMod),
            data      = kims,
            group     = "sex",
            ordered   = TRUE,
            estimator = "WLSMV")

## Summarize the results:
summary(tOut)

## Check the model fit:
fitMeasures(tOut, stats)
```

As expected, the thresholds are equal across groups. All the parameter estimates seem sensible, and the model still fits very well.

---

##

**Test threshold invariance.**

```{r}
## Test threshold invariance via model comparisons:
compareFit(baseOut, tOut) |> summary()
```

The $\Delta \chi^2$ test is nonsignificant, so we can infer invariant thresholds.

---

##

**Define the model syntax for the weak/metric invariance model.**

- Use the `measEq.syntax()` function from the **semTools** package to generate the model syntax.
- Set the scale by standardizing the latent factors (i.e., use the fixed-factor/reference-group method of identification).
- Use the delta parameterization to define the latent response variables.
- Use the Wu & Estabrook (2016) identification conditions.

```{r}
## Define the model syntax for the threshold/loading-invariant model:
tlMod <- measEq.syntax(configural.model = cfaMod,
                       data             = kims,
                       ordered          = TRUE,
                       parameterization = "delta",
                       ID.fac           = "std.lv",
                       ID.cat           = "Wu.Estabrook.2016",
                       group            = "sex",
                       group.equal      = c("thresholds", "loadings")
                       )

## Check the resulting syntax:
tlMod |> as.character() |> cat()
```

---

##

**Estimate the weak invariant model.**

- Check the results by summarizing the model and evaluating the model fit.

```{r}
## Estimate the threshold/loading-invariant model:
tlOut <- cfa(model     = as.character(tlMod),
             data      = kims,
             group     = "sex",
             ordered   = TRUE,
             estimator = "WLSMV",
             int.ov.free = TRUE)

## Summarize the results:
summary(tlOut)

## Check the model fit:
fitMeasures(tlOut, stats)
```

Again, model fit is still excellent, and all the parameter estimates are reasonable.

---

##

**Test weak invariance.**

```{r}
## Test threshold/loading invariance via model comparisons:
compareFit(tOut, tlOut) |> summary()
```

Relative to the threshold-invariant model, we haven't lost a significant amount of fit by equating the factor loadings
across groups. Conditioning on the dubious threshold invariance, we can claim weak invariance.

---

##

**Define the model syntax for the strong/scalar invariance model.**

- Use the `measEq.syntax()` function from the **semTools** package to generate the model syntax.
- Set the scale by standardizing the latent factors (i.e., use the fixed-factor/reference-group method of identification).
- Use the delta parameterization to define the latent response variables.
- Use the Wu & Estabrook (2016) identification conditions.

```{r}
## Define the model syntax for the threshold/loading/intercept-invariant model:
tliMod <-
    measEq.syntax(configural.model = cfaMod,
                  data             = kims,
                  ordered          = TRUE,
                  parameterization = "delta",
                  ID.fac           = "std.lv",
                  ID.cat           = "Wu.Estabrook.2016",
                  group            = "sex",
                  group.equal      = c("thresholds", "loadings", "intercepts")
                  )

## Check the syntax:
tliMod |> as.character() |> cat()
```

---

##

**Estimate the weak invariant model.**

- Check the results by summarizing the model and evaluating the model fit.

```{r}
## Estimate the threshold/loading/intercept-invariant model:
tliOut <- cfa(model     = as.character(tliMod),
              data      = kims,
              group     = "sex",
              ordered   = TRUE,
              estimator = "WLSMV")

## Summarize the results:
summary(tliOut)

## Check the model fit:
fitMeasures(tliOut, stats)
```

The strongly invariant model still fits very well.

```{r}
## Test threshold/loading/intercept invariance via model comparisons:
compareFit(tlOut, tliOut) |> summary()
```

Now, we lose a significant amount of fit by constraining the item intercepts in addition to the factor loadings and
thresholds (i.e., the $\Delta \chi^2$ test is significant). So, we should reject the hypothesis of strong invariance.

---

# Moderated SEM

---

```{r}
semMod <- paste(
  cfaMod,
  'acc ~ c("b1f", "b1m") * des + c("b2f", "b2m") * obs + c("b3f", "b3m") * age',
  'age ~~ des + obs',
  sep = "\n"
)

modOut <- sem(
  model = semMod,
  data = kims,
  group = "sex",
  ordered = cats,
  estimator = "WLSMV",
  parameterization = "delta",
  std.lv = TRUE,
  group.equal = c("thresholds", "loadings")
)

summary(modOut)
```

```{r}
cons <- c('b1f == b1m', 'b2f == b2m', 'b3f == b3m')

lavTestWald(modOut, constraints = cons)
```

```{r}
semMod <- paste(
  cfaMod,
  'acc ~ c("b1", "b1") * des + c("b2", "b2") * obs + c("b3", "b3") * age',
  'age ~~ des + obs',
  sep = "\n"
)

# tlSemMod <-
#     measEq.syntax(configural.model = semMod,
#                   data             = kims,
#                   ordered          = cats,
#                   parameterization = "delta",
#                   ID.fac           = "std.lv",
#                   ID.cat           = "Wu.Estabrook.2016",
#                   group            = "sex",
#                   group.equal      = c("thresholds", "loadings")
#                   )

modOut2 <- sem(
  model = semMod,
  data = kims,
  group = "sex",
  ordered = cats,
  estimator = "WLSMV",
  parameterization = "delta",
  std.lv = TRUE,
  group.equal = c("thresholds", "loadings")
)

summary(modOut)
```

```{r}
lavTestLRT(modOut, modOut2)
anova(modOut, modOut2)
```

---

End of Lab 8

---
