---
title: "Lab: Model Selection"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

Today’s lab meeting consists of three exercises. Each exercise illustrates the evaluation of theory-based hypotheses in a SEM model (using the goric() function of the restriktor package). 

The exercises are based on the following article:

Kuiper, R. M. (2022). AIC-type Theory-Based Model Selection for Structural Equation Models, Structural Equation Modeling: A Multidisciplinary Journal, 29:1, 151-158
[https://doi-org.proxy.library.uu.nl/10.1080/10705511.2020.1836967].

&nbsp;

Practical information:

- All the data files for these exercises can be found at the course platform (i.e., Ulearning platform). Make sure to download and unzip the .zip file. The folder containing the data files will probably be your working directory.


Short description of exercises:

- Exercise 1 addresses theory-based hypotheses in a two-factor confirmatory factor analysis,

- Exercise 2 addresses theory-based hypotheses in a latent regression and,

- Exercise 3 addresses theory-based hypotheses in a multigroup latent regression.





## Data

The exercises below will all be applied to the same data set: a simulated data set based on the Sesame Street data (Stevens, 1996), which is included as the dataset ‘sesamsim’ in the gorica package. Thus, when you install and load the gorica package, the object sesamsim exists, which contains all the data:
```{r, warning=F, message=F}
# Install and load the gorica library because it contains the data 'sesamesim'
if (!require("gorica")) install.packages("gorica") # install this package first (once)
library(gorica)
# 
# If you want to peak at the top of the data set
#head(sesamesim)
```


Since the data sets contains simulated data, there are no missing values. All the variables are numeric, which does not affect the results in these exercises since there are no grouping variables with more than two level. We need only one grouping variable, namely sex, which consists of two levels. It can be helpful, though, to make it a factor such that labels can be assigned to the two levels:

```{r, warning=F, message=F, eval=F}
sesamesim$sex <- factor(sesamesim$sex, labels = c("boy", "girl"))
```

&nbsp;

The exercises concern the effect of watching one year of the tv-series “Sesame Street” on the knowledge of numbers of N = 240 children aged between 34 to 69 months. Several variables have been measured before and after watching Sesame Street for one year: Knowledge of numbers before (Bn) and after (An) watching, and analogously, knowledge of body parts (Bb and Ab), letters (Bl and Al), forms (Bf and Af), relationships (Br and Ar), and classifications (Bc and Ac). The score ranges on these variables ranges from ‘1 to 20’ to ’1 to 70‘. In the exercises, we will use these variables as well as the following ones: biological age in months (age; score range: 34 to 69), the Peabody test measuring the mental age of children (peabody; score range: 15 to 89), and gender (sex; 1 = boy, 2 = girl).


## Exercise 1: Confirmatory Factor Analysis

In this exercise, we will inspect the two-factor confirmatory factor analysis (CFA), in which the A(fter) measurements (i.e., Ab, Al, Af, An, Ar, and Ac) load on the factor A, and the B(efore) measurements load on the factor B.

### 1a 

It is a good practice to draw the model you plan to estimate for yourself. Draw the model.


<details><summary><b>Click to show answers</b></summary>

&nbsp;

You can compare your path diagram to Figure 1 in https://doi-org.proxy.library.uu.nl/10.1080/10705511.2020.1836967 or compare it to one created based on the fitted object in Exercise 1d.

</details>

&nbsp;


### 1b 

Specify the two-factor CFA model (such that it can be used in the lavaan function later on).

For ease of specifying the hypotheses (done next), label the parameters; e.g. use for the factor Ab: A1*Ab

*Make sure* to install and load all the packages needed (lavaan, restriktor, and gorica).


<details><summary><b>Click to show answers</b></summary>

&nbsp;

**Installing and loading packages**

```{r, warning=F, message=F}
# install this package first (once)
if (!require("restriktor")) install.packages("restriktor")
if (!require("lavaan")) install.packages("lavaan")

# Load
library(restriktor)
library(lavaan)

# Also install and load the gorica library because it contains the data 'sesamesim'
if (!require("gorica")) install.packages("gorica") # install this package first (once)
library(gorica)
```

**Specify the CFA model**

Note: the `goric` function cannot always use the default labeling, so give your own labels to estimates by including them in the `lavaan` model:

```{r, warning=F, message=F}
model1 <- '
    A =~ A1*Ab + A2*Al + A3*Af + A4*An + A5*Ar + A6*Ac 
    B =~ B1*Bb + B2*Bl + B3*Bf + B4*Bn + B5*Br + B6*Bc 
'
```

</details>

&nbsp;


### 1c 

Come up with one or more hypotheses and a failsafe hypothesis.

Remarks:

* In a paper, hypotheses are specified in terms of population parameters (often expressed as Greek letters).
* In R, hypotheses are specified in terms of the model parameters, which will here be our assigned labels.
* When using goric(), the restrictions need to be 'connected' by using ';', e.g.: A1 > .6; A2 > .6.
* Make sure that the comparison of factor loadings or comparison of factor loadings to a pre-specified value is fair. 
  *Hint:* You probably have hypotheses w.r.t. standardized model parameters.
* In the solution below, one theory-based hypothesis is evaluated against its complement. The hypothesis states that each standardized factor loading is larger than .6, meaning that the indicators are strongly related to the factors to which they are assigned. 


<details><summary><b>Click to show answers</b></summary>

&nbsp;

**Hypothesis of interest**

In this solution, we hypothesize that the indicators are strongly related to the factors to which they are assigned. This can be reflected by stating that each standardized factor loading is larger than .6. This leads to 12 order restrictions, which should be 'connected' by using ';'.

The hypothesis of interest (using our own labeling) is then formulated by:

```{r, warning=F, message=F}
H1.1 <- "
A1 > .6; A2 > .6; A3 > .6; A4 > .6; A5 > .6; A6 > .6; 
B1 > .6; B2 > .6; B3 > .6; B4 > .6; B5 > .6; B6 > .6
"
```

</details>

&nbsp;


### 1d 

Run the two-factor CFA specified in 1b.


<details><summary><b>Click to show answers</b></summary>

&nbsp;

**Fit the CFA model**

```{r, warning=F, message=F}
fit1 <- sem(model1, data = sesamesim, std.lv = TRUE)

# summary(fit1, standardize = T)
# fitmeasures(fit1)
```

**Plot the CFA model**
```{r, warning=F, message=F}
# install this package first (once)
if (!require("lavaanPlot")) install.packages("lavaanPlot") 
library(lavaanPlot)

# plot the model
lavaanPlot(model = fit1, node_options = list(shape = "box", fontname = "Helvetica"), 
           edge_options = list(color = "grey"), coefs = T, stand = T, covs = T)
```

</details>

&nbsp;


### 1e 

Use the goric() function in the restriktor package to evaluate your hypothesis/-es of interest with the GORICA (`type = "gorica"`).

* In case there is one hypothesis of interest, then your failsafe or competing hypothesis is its complement. Then, use: `comparison = "complement"`
* In case the hypotheses address standardized model parameters, also include the argument: `standardized = TRUE`
* Before running goric(), set a seed value (e.g., `set.seed(100)`), since the penalty (PT) is obtained via sampling. One can also do a sensitivity check by using other seed values.

*Interpret the output. What are your conclusions?*


<details><summary><b>Click to show answers</b></summary>

&nbsp;

**Apply GORICA**

```{r, message=F, warning=FALSE}
set.seed(100) # since penalty (PT) is obtained via sampling.
results1 <- goric(fit1, H1.1, comparison = "complement", type = "gorica", 
                    standardized = TRUE) 
```

**Summary GORICA**

In this case, there are 12 order restrictions and 12 parameters. Then, hus the computation time is somewhat long. See below for a possibility to improve the computation time.

```{r, message=F, warning=FALSE}
# Note: This also includes the comparison of hypotheses
summary(results1)
```

The order-restricted hypothesis ‘H1.1’ has 85.561 times more support than its complement.

**Possible conclusion**

* The R output shows that the hypothesis of interest (H1.1) has the largest fit and the smallest complexity and, thus, the smallest GORICA value and highest GORICA weight. 
* The GORICA weight for H1 (against it complement) is 0.988.
* Hence, the support in the data in favor of H1 is overwhelming: H1.1 is $0.988/0.012 \approx 86$ times more supported than its complement.
* Thus, there is overwhelming support for the hypothesis that each factor loading is larger than .6.

</details>

&nbsp;


<details><summary><b>Click to show code with alternative PT calculation</b></summary>

&nbsp;

**Alternative PT calculation**

The default way of calculating the PT in `goric()` is slow(er) when the number of parameters is large and the number of restrictions equal the number of parameters. In contrast, when the number of restrictions is not large or when the number of restrictions is smaller that the number of parameters, the default method is fast.

Here, there are 12 parameters and 12 restrictions, so may want to use: `mix.weights = "boot"`. That is, we will use bootstrap in the calculation of the level probabilities (LPs) needed in PT. The results of course do not change, but the computation time may. 

```{r, message=F, warning=FALSE}
# install this package first (once)
if (!require("parallel")) install.packages("parallel")
library(parallel)

# Determine number of cores that can be used, to fasten the calculation of PT even more.
nrCPUcores <- detectCores(all.tests = FALSE, logical = TRUE)

set.seed(100)
results1_b <- goric(fit1, H1.1, comparison = "complement", type = "gorica", standardized = TRUE, 
                   mix.weights = "boot", parallel = "snow", ncpus = nrCPUcores, mix.bootstrap = 99999)
summary(results1_b) # Note: This also includes the comparison of hypotheses
```

</details>

&nbsp;


<details><summary><b>Click to show code for a sensitivity check</b></summary>

&nbsp;

**Sensitivity check**

Although the penalty (PT) calculated by the goric() seems to be very robust, it cannot harm to do a sensitivity check on the PT when you use the goric(). Next, you can see that the influence of the seed on the PT values is negligible, as we would also like to see.

```{r, message=F, warning=FALSE}
set.seed(100)
goric(fit1, H1.1, comparison = "complement", type = "gorica", standardized = TRUE, 
      mix.weights = "boot", parallel = "snow", ncpus = nrCPUcores, mix.bootstrap = 99999)$result[,3]

results1_b$result[,3]

set.seed(100100)
goric(fit1, H1.1, comparison = "complement", type = "gorica", standardized = TRUE, 
       mix.weights = "boot", parallel = "snow", ncpus = nrCPUcores, mix.bootstrap = 99999)$result[,3]

set.seed(123456)
goric(fit1, H1.1, comparison = "complement", type = "gorica", standardized = TRUE, 
       mix.weights = "boot", parallel = "snow", ncpus = nrCPUcores, mix.bootstrap = 99999)$result[,3]
```

</details>

&nbsp;


### Exercise 2: Latent regression

In this exercise, we will inspect a latent regression model. The factors B and A have the same indicators as in Exercise 1. The difference is the addition of a latent regression in which A is regressed on B, age, and peabody, to investigate whether children’s knowledge after watching Sesame Street for a year is predicted by their knowledge one year before, as well as by their biological and mental age, which have a correlation of 
```{r, message=F, warning=FALSE}
cor(sesamesim$peabody, sesamesim$age)
```
($\approx .24$), so there is no multicollinearity.

Repeat the steps in Exercise 1, but now for this model, summarized by:


### 2a 

Draw the model.


<details><summary><b>Click to show answers</b></summary>

&nbsp;

You can compare your path diagram to Figure 2 in https://doi-org.proxy.library.uu.nl/10.1080/10705511.2020.1836967 or compare it to one created based on the fitted object in Exercise 2d.

</details>

&nbsp;


### 2b 

Specify the latent regression model, using labels.


<details><summary><b>Click to show answers</b></summary>

&nbsp;

**Installing and loading packages**

```{r, warning=F, message=F}
# install this package first (once)
if (!require("restriktor")) install.packages("restriktor")
if (!require("lavaan")) install.packages("lavaan")

# Load
library(restriktor)
library(lavaan)

# Also install and load the gorica library because it contains the data 'sesamesim'
if (!require("gorica")) install.packages("gorica") # install this package first (once)
library(gorica)
```


**Specify the latent regression model**

```{r, message=F, warning=FALSE}
model2 <- '
    A =~ Ab + Al + Af + An + Ar + Ac 
    B =~ Bb + Bl + Bf + Bn + Br + Bc 
    A ~ AB*B + AAge*age + APeabody*peabody
'
```

</details>

&nbsp;


### 2c 

Come up with one or more hypotheses and a failsafe hypothesis.


<details><summary><b>Click to show information on the hypotheses used in the answers</b></summary>

&nbsp;

In the solution, we evaluate three theory-based hypotheses (H1-H3) on standardized regression coefficients. Since these three do not overlap all the possible theories, we include the unconstrained as failsafe.

* **H1** specifies that a larger score on B corresponds to a larger score on A (i.e., a positive relation between B and A) and that age and peabody do not predict A.

* **H2** specifies that the positive relation between B and A is stronger than the positive relation between peabody and A and that age cannot be used to predict A.

* **H3** specifies that the predictive power of B is larger than that of peabody, which, in turn, is larger than that of age which in turn is positive. 

Notably, only in case all these hypotheses are of interest, these should all be included in the set; especially if there is overlap like here (see https://doi-org.proxy.library.uu.nl/10.1080/10705511.2020.1836967 for more information).

</details>

&nbsp;


<details><summary><b>Click to show answers</b></summary>

&nbsp;

**Specify hypothesis of interest**

```{r, message=F, warning=FALSE}
H1.2 <- "AB > APeabody = AAge = 0"
H2.2 <- "AB > APeabody > AAge = 0" 
H3.2 <- "AB > APeabody > AAge > 0"
```

</details>

&nbsp;


### 2d 

Run the latent regression model.


<details><summary><b>Click to show answers</b></summary>

&nbsp;

**Fit the latent regression model**

Fit the latent regression model using the `lavaan::sem()` (or `lavaan()`) function

```{r, message=F, warning=FALSE}
fit2 <- sem(model2, data = sesamesim, std.lv = TRUE)
```

**Plot the latent regression model**

```{r, message=F, warning=FALSE}
# install this package first (once)
if (!require("lavaanPlot")) install.packages("lavaanPlot") 
library(lavaanPlot)

# plot
lavaanPlot(model = fit2, node_options = list(shape = "box", fontname = "Helvetica"), 
           edge_options = list(color = "grey"), coefs = T, stand = T, covs = T)

# Alternative
# install this package first (once)
if (!require("tidySEM")) install.packages("tidySEM") 
library(tidySEM)

# plot
graph_sem(fit2)

# Another alternative
# install this package first (once)
if (!require("semPlot")) install.packages("semPlot") 
library(semPlot)

# plot
semPaths(fit2, "par", weighted = FALSE, nCharNodes = 7, shapeMan = "rectangle",
         sizeMan = 8, sizeMan2 = 5)
```

</details>

&nbsp;


### 2e 

Evaluate the hypotheses with the GORICA.

*Interpret the output. What are your conclusions? *


<details><summary><b>Click to show answers</b></summary>

&nbsp;

**GORICA**

In our hypotheses (as in many not addressing mean values), we need standardized estimates for a meaningful comparison. Because we have more than one hypothesis, we need to use the unconstrained as failsafe, which is the default option in goric(). This is requested via:

```{r, message=F, warning=FALSE}
set.seed(100)
results2 <- goric(fit2, H1.2, H2.2, H3.2, type = "gorica", standardized = TRUE) 
summary(results2) # Note: This also includes the comparison of hypotheses
```

Note: In case of equal log-likelihood (loglik) values, the ratio weights are solely based on the difference in penalty values (see https://doi-org.proxy.library.uu.nl/10.1080/10705511.2020.1836967 for more information).

**Possible conclusions**

* Since the hypotheses overlap and the most restricted one (H1.2) receives the most support (and is not weak), H1.2 is the preferred hypothesis. 
* Because the support for the other hypotheses also contain support for H1.2, one could also compare H1.2 to its complement, see below. This shows convincing support for H1.2. 
* Thus, there is support for the hypothesis that a larger score on B corresponds to a larger score on A (i.e., a positive effect) and that age and peabody do not predict A. Based on the evaluation below, it is approximately 7 times more likely than its complement containing competing hypotheses.


**Extra: H1.2 vs its complement**

Since the hypotheses are nested, the hypotheses share support (see https://doi-org.proxy.library.uu.nl/10.1080/10705511.2020.1836967 for more information). Therefore, we also examine the best of these (i.e., H1.2) against its compliment:

```{r, message=F, warning=FALSE}
set.seed(100)
results2_c <- goric(fit2, H1.2, comparison = "complement", type = "gorica", 
                    standardized = TRUE) 
summary(results2_c) # Note: This also includes the comparison of hypotheses
#
#        model  loglik  penalty   gorica  gorica.weights
#1        H1.2   6.836    0.500  -12.672           0.874
#2  complement   6.894    2.500   -8.789           0.126
```

The order-restricted hypothesis ‘H1.2’ has  6.967 times more support than its complement. This is convincing support for H1.2.

</details>

&nbsp;


<details><summary><b>Click to show code for a sensitivity check</b></summary>

&nbsp;

**Sensitivity check**

Next, you find code to check the influence of the seed in calculating the PT, which turns out to be negligible.

```{r, message=F, warning=FALSE}
set.seed(100)
goric(fit2, H1.2, H2.2, H3.2, type = "gorica", standardized = TRUE)$result[,3]
results2$result[,3]

set.seed(100100)
goric(fit2, H1.2, H2.2, H3.2, type = "gorica", standardized = TRUE)$result[,3]

set.seed(123456)
goric(fit2, H1.2, H2.2, H3.2, type = "gorica", standardized = TRUE)$result[,3]
```

</details>

&nbsp;


### Exercise 3: Multigroup latent regression

In this exercise, we will inspect a multi-group regression model, by including the grouping variable gender (sex) in a regression model. This means that there is one regression model for girls and one for boys, where the standardized model parameter estimates may differ between girls and boys. 

In the regression, postnumb is regressed on prenumb, to investigate whether children’s knowledge of numbers after watching Sesame Street for a year is predicted by their knowledge of numbers one year before. 

Repeat the steps in Exercise 1, but now for this model, summarized by:


### 3a 

Draw the model.


<details><summary><b>Click to show answers</b></summary>

&nbsp;

You can compare your path diagram to the one created based on the fitted object in Exercise 3d.

</details>

&nbsp;


### 3b 

Specify the multigroup regression model, using labels.


<details><summary><b>Click to show answers</b></summary>

&nbsp;

**Installing and loading packages**

Load the `restriktor` and `lavaan` libraries:

```{r, warning=F, message=F, eval=F}
# install this package first (once)
if (!require("restriktor")) install.packages("restriktor") 
# install this package first (once)
if (!require("lavaan")) install.packages("lavaan") 

library(restriktor)
library(lavaan)

# Also load the gorica library because it contains the data 'sesamesim'
# install this package first (once)
if (!require("gorica")) install.packages("gorica") 
library(gorica)
```

**Data**

Optional: Make the variable 'sex' a factor with the right labels:
```{r, warning=F, message=F}
sesamesim$sex <- factor(sesamesim$sex, labels = c("boy", "girl"))
```

**Specify the multigroup regression model**

Specify the multigroup regression model, using own labels:

```{r, warning=F, message=F}
model3 <- '
    postnumb ~ c(Pre_b, Pre_g)*prenumb 
'
```

</details>

&nbsp;


### 3c

Come up with one or more hypotheses and a failsafe hypothesis.


<details><summary><b>Click to show answers</b></summary>

&nbsp;

**Hypothesis of interest**

In the solution, we evaluate one theory-based hypothesis against its complement. The hypothesis states that the standardized relationship between postnumb and prenumb is higher for girls than for boys. In this case, the complement states that the standardized relationship between postnumb and prenumb is lower for girls than for boys.

The hypotheses of interest is formulated as follows, using our own labeling:

```{r, warning=F, message=F}
H1.3 <- "Pre_b < Pre_g"
```

</details>

&nbsp;


### 3d

Run the multigroup regression model.


<details><summary><b>Click to show answers</b></summary>

&nbsp;

**Fit the multigroup regression model**

Fit the multiple group regression model using the sem() function in lavaan:

```{r, warning=F, message=F}
fit3 <- sem(model3, data = sesamesim, std.lv = TRUE, group = "sex")
```

**Plot the multigroup regression model**

```{r, warning=F, message=F}
# install this package first (once)
#if (!require("lavaanPlot")) install.packages("lavaanPlot") 
#library(lavaanPlot)

# plot
lavaanPlot(model = fit3, node_options = list(shape = "box", fontname = "Helvetica"),
           edge_options = list(color = "grey"), coefs = T, stand = T, covs = T)
```

</details>

&nbsp;


### 3e 

Evaluate the hypotheses with the GORICA.

*Interpret the output. What are your conclusions? *


<details><summary><b>Click to show answers</b></summary>

&nbsp;

**GORICA**

```{r, warning=F, message=F}
set.seed(100)
results3 <- goric(fit3, H1.3, comparison = "complement", type = "gorica", 
                  standardized = T) 

summary(results3) # Note: This also includes the comparison of hypotheses
```


**Possible conclusions**

* The output shows that the hypothesis of interest and its complement are equally likely, since both have a weight of approximately .50. 
  That is, they have about the same support.
* Since the hypotheses do not overlap and are equally complex (i.e., have the same penalty value), this implies that their boundary is the preferred hypothesis, that is, H0: Pre_b = Pre_g.
* Thus, there is support for the boundary of the hypothesis of interest and its complement, indicating that the relationship between postnumb and prenumb is equally high for girls and boys.

See https://doi-org.proxy.library.uu.nl/10.1080/10705511.2020.1836967 for more information.


</details>

&nbsp;


<details><summary><b>Click to show code for a sensitivity check</b></summary>

&nbsp;

**Sensitivity check**

Next, you find code to check the influence of the seed in calculating the PT, which is none:

```{r, warning=F, message=F}
set.seed(100)
goric(fit3, H1.3, comparison = "complement", 
      type = "gorica", standardized = T)$result[,3]

results3$result[,3]

set.seed(100100)
goric(fit3, H1.3, comparison = "complement", 
      type = "gorica", standardized = T)$result[,3]

set.seed(123456)
goric(fit3, H1.3, comparison = "complement", 
      type = "gorica", standardized = T)$result[,3]
```

</details>

&nbsp;


### 3f - Extra

Do the same evaluation as in 3b to 3e, but now use the (standardized) estimates and their covariance matrix as input to the goric function (instead of the lavaan object). E.g., use:

`goric(est, VCOV = vcov, H1, comparison = "complement", type = "gorica ")`

with `est` the (standardized) estimates and `vcov` their covariance matrix.



<details><summary><b>Click to show answers</b></summary>

&nbsp;

**Installing and loading packages**

```{r, warning=F, message=F, eval=F}
# install this package first (once)
if (!require("restriktor")) install.packages("restriktor") 
# install this package first (once)
if (!require("lavaan")) install.packages("lavaan") 

library(restriktor)
library(lavaan)

# Also load the gorica library because it contains the data 'sesamesim'
# install this package first (once)
if (!require("gorica")) install.packages("gorica") 
library(gorica)
```

**Data**

Optional: Make the variable 'sex' a factor with the right labels:
```{r, warning=F, message=F}
sesamesim$sex <- factor(sesamesim$sex, labels = c("boy", "girl"))
```

**Specify the multigroup regression model**

```{r, warning=F, message=F}
model3 <- '
    postnumb ~ prenumb 
'
```

**Hypothesis of interest**

```{r, warning=F, message=F}
H1.3 <- "Pre_boy < Pre_girl"
```
We, again, used our own labeling, which should be the same as those used below:


**Fit the multigroup regression model**

```{r, warning=F, message=F}
fit3 <- sem(model3, data = sesamesim, std.lv = TRUE, group = "sex")
```


**Plot the multigroup regression model**

```{r, warning=F, message=F}
# install this package first (once)
#if (!require("lavaanPlot")) install.packages("lavaanPlot")
library(lavaanPlot)

#plot 
lavaanPlot(model = fit3, node_options = list(shape = "box", fontname = "Helvetica"), edge_options = list(color = "grey"), coefs = T, stand = T, covs = T)
```

**GORICA**

Now, we will extract the estimates from the lavaan object ourselves and use that.

Obtain standardized(!) estimates from lavaan object (labelled or unlabelled) via:

```{r, warning=F, message=F}
est_3 <- lavaan::standardizedsolution(fit3)[c(1,6), 'est.std']
# Note: This should be the same as the labeling used the H1.3.
names(est_3) <- c("Pre_boy", "Pre_girl") 

# Note: Use this in the 'VCOV = vcov_3' command.
vcov_3 <- lavInspect(fit3, "vcov.std.all")[c(1,4), c(1,4)] 
```

Now, use these when calculating the GORICA values and weights for H1.3 and its complement (`comparison = "complement"`), via:

```{r, warning=F, message=F}
set.seed(100)
results3 <- goric(est_3, VCOV = vcov_3, H1.3, comparison = "complement", type = "gorica") 
summary(results3) # Note: This also includes the comparison of hypotheses
```


**Possible conclusions**

* The output shows that the hypothesis of interest and its compliment are equally likely, since both have a weight of approximately .50. That is, they have about the same support.
* Since the hypotheses do not overlap and are equally complex (i.e., have the same penalty value), this implies that their boundary is the preferred hypothesis, that is, H0: Pre_boy = Pre_girl.
* Thus, there is support for the boundary of the hypothesis of interest and its complement, indicating that the relationship between postnumb and prenumb is equally high for girls and boys.


</details>

&nbsp;


<details><summary><b>Click to show code for a sensitivity check</b></summary>

&nbsp;

**Sensitivity check**

Next, you find code to check the influence of the seed in calculating the PT, which is none:

```{r, warning=F, message=F}
set.seed(100)
goric(est_3, VCOV = vcov_3, H1.3, comparison = "complement", 
      type = "gorica")$result[,3]

results3$result[,3]

set.seed(100100)
goric(est_3, VCOV = vcov_3, H1.3, comparison = "complement", 
      type = "gorica")$result[,3]

set.seed(123456)
goric(est_3, VCOV = vcov_3, H1.3, comparison = "complement", 
      type = "gorica")$result[,3]
```

</details>

&nbsp;



